Accelerated oblique random survival forests

### 2022-07-16 paper/jmlr/main.tex
paper v1

This study makes two contributions to oblique RSFs. First, we reduce their computational cost (\ie, accelerate them) with a scalable algorithm to identify linear combinations of coefficients. In a general benchmark experiment including 31 risk prediction tasks, we show that accelerated oblique RSFs are roughly 500 times faster with equivalent or superior prediction accuracy compared to existing routines to fit oblique RSFs. Second, we improve their interpretability with `negation VI', a method to estimate VI that engages with coefficients in linear combinations of predictors instead of permuting predictor values. In simulation studies where VI is estimate using permutation, analysis of variance (ANOVA; see \citet{menze2011oblique}), and approximations of Shapley values, we find that negation VI improves the oblique RSF's ability to discriminate between correlated signal and noise variables. The accelerated oblique RSF and multiple methods to compute VI for oblique RSFs (permutation, ANOVA, and negation) are available in the \texttt{aorsf} R Package.

@@

@@

We use a collection of 18 publicly available data sets to benchmark the prediction accuracy and computational efficiency of the accelerated ORSF and each of the other learners described in \secref{sec:learners}. The number of right-censored outcomes per data set ranged from one to four, and the total number of risk prediction tasks we analyzed was 31 (\tabrefAppendix{tab:datasets}). Across all prediction tasks, the number of observations ranged from 137 to 17,549 (median: 2,231), the number of predictors ranged from 7 to 1,692 (median: 41), and the percentage of censored observations ranged from 5.26 to 97.7 (median: 78.2).

@@

For each risk prediction task, we completed 20 runs of Monte-Carlo cross validation. In each run, we used a random sample containing 50\% of the available data for training and the remaining 50\% for testing each of the learners described in \secref{sec:learners}. Then, for each learner, we computed the IPA, time-dependent C-statistic, and computational time required to fit a prediction model and compute risk predictions. If any learner failed to obtain predictions on any particular split of data\footnote{For example, when the prediction task was to predict risk of death in the ACTG 320 clinical trial (26 events total), some splits did not leave enough events in the training data to fit complex learners such as the neural network}, the results for that split were omitted from downstream analyses.

@@

After collecting data from 20 replications of Monte-Carlo cross validation for the 14 learners in all 31 risk prediction tasks, we analyzed the resulting 8,680 observations of IPA and, separately, time-dependent C-statistic, using a Bayesian linear mixed model. Our approach follows the ideas described by \citet{benavoli2017time} and \citet{tidymodels}, who developed guidelines on making statistical comparisons between learners using Bayesian models. Specifically, we fit two models: $$\text{IPA} = \widehat{\gamma}_0 + \widehat{\gamma} \cdot \text{learner} + (1\,|\, \text{data/run}) $$ and $$\text{C-stat} = \widehat{\gamma}_0 + \widehat{\gamma} \cdot \text{learner} + (1\,|\, \text{data/run}).$$ Random intercepts for specific splits of data (\ie, \texttt{run} in the model formula) were nested within datasets. The intercept, $\widehat{\gamma}_0$, was the expected value of the outcome using \texttt{aorsf-fast}, making the coefficients in $\widehat{\gamma}$ the expected differences between \texttt{aorsf-fast} and other learners. Default priors from \texttt{rstanarm} were applied for model fitting \citep{rstanarm}.

\paragraph{Hypothesis testing} For both the IPA and time-dependent C-statistic, we conducted equivalence and inferiority tests based on a 1 point region of practical equivalence. More specifically, we concluded that two learners had practically equivalent IPA or time-dependent C-statistic if there was a 95\% or higher posterior probability that the absolute difference in the relevant metric was less than 1. We concluded that one learner was weakly superior when there was $\geq$ 0.95\% posterior probability that the difference in the relevant metric was non-zero, and concluded superiority when when there was $\geq$ 0.95\% posterior probability that the difference in the relevant metric was 1 or more.

@@

Compared to learners that were not oblique RSFs, \texttt{aorsf-fast} had the highest IPA in  out of 31 risk prediction tasks, with an overall mean IPA of  (Figure \ref{fig:bm_pred_viz_ibs}). Compared to the learner with the second highest mean IPA (\texttt{rsf-standard}), \texttt{aorsf-fast}'s mean was 1.27 points higher, a relative increase of 10.5\%. The posterior probability of \texttt{aorsf-fast} and \texttt{aorsf-cph} having practically equivalent expected IPA was 0.98, and the posterior probability of \texttt{aorsf-fast} having a superior IPA to other learners ranged from 0.76 (versus \texttt{rsf-standard}) to $>$0.999 (versus several other learners; see Figure \ref{fig:bm_pred_model_viz_ibs})

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/bm_pred_viz_ibs-1} \caption[Index of prediction accuracy for the accelerated oblique random survival forest and other learning algorithms across multiple risk prediction tasks]{Index of prediction accuracy for the accelerated oblique random survival forest and other learning algorithms across multiple risk prediction tasks. Text appears in tasks where the accelerated oblique random survival forest obtained the highest index of prediction accuracy, showing the absolute and percent improvement over the second best learner.}\label{fig:bm_pred_viz_ibs}
\end{figure}

@@

Compared to learners that were not oblique RSFs, \texttt{aorsf-fast} had the highest time-dependent C-statistic in  out of 31 risk prediction tasks, with an overall mean of  (Figure \ref{fig:bm_pred_viz_cstat}). Compared to the learner with the second highest mean C-statistic (\texttt{obliqueRSF-net}), \texttt{aorsf-fast}'s mean was 0.643 points higher, a relative increase of 0.837\%. The posterior probability of \texttt{aorsf-fast} and \texttt{aorsf-cph} having practically equivalent expected time-dependent C-statistics was 0.99, and the posterior probability of \texttt{aorsf-fast} having a superior time-dependent C-statistic versus other learners ranged from 0.01 (versus \texttt{obliqueRSF-net}) to $>$0.999 (versus several other learners; see Figure \ref{fig:bm_pred_model_viz_cstat})

@@

Overall, \texttt{aorsf-fast} was the second fastest learner, with an expected model development and risk prediction time about 1/2 second longer than \texttt{glmnet-cox} (Figure \ref{fig:bm_pred_time}). Compared to its predecessor, \texttt{obliqueRSF-net}, \texttt{aorsf-fast} was XYZ times faster.

@@

Parameters that varied in the current simulation study included the number of observations (1000, 3000, and 5000) and the minimum and maximum correlation between predictors (-0.1 to 0.1, -0.3 to 0.3, and -0.5 to 0.5). Parameters that remain fixed throughout the study included the number of predictors in each class (15) and the effect size of each predictor, with an increase of one standard deviation associated with a 64\% increase in relative risk.

@@

Overall, \texttt{aorsf-negate} had the highest C-statistic in 27 out of 36 VI tasks, with an overall mean of 76.1 (Figure \ref{fig:bm_vi_viz}). Compared to the VI technique with the second highest overall C-statistic (\texttt{aorsf-anova}), \texttt{aorsf-negate}'s mean was 2.22 points higher, a relative increase of 3.00\%. Among the four relevant variable classes, \texttt{aorsf-negate} had the highest mean C-statistic for main effects, combination effects, and non-linear effects, with the greatest advantage of using \texttt{aorsf-negate} occurring among non-linear and combination variables.

The three techniques that used `aorsf` to estimate VI were ranked first (negation), second (ANOVA), and third (permutation) in overall mean C-statistic across all of the simulation scenarios. Computationally, ANOVA VI was the fastest technique with a median average

@@

In this paper, we have developed two contributions to the oblique RSF: (1) the accelerated oblique RSF (\ie, \texttt{aorsf-fast}) and (2) negation VI. Our technique to accelerate the oblique RSF reduces the number of operations required to find linear combinations of inputs using a single iteration of Newton Raphson scoring, while our VI technique directly engages with coefficients in linear combinations of inputs to measure importance of individual variables. In numeric experiments, we found that that \texttt{aorsf-fast} is orders of magnitude more efficient and just as accurate in risk prediction tasks compared to its predecessor, \texttt{obliqueRSF-net}. We also found several cases where negation VI allows oblique RSF models to discriminate between relevant and irrelevant variables more effectively than three standard methods to estimate VI: permutation, ANOVA, and SHAP VI. Our results favored negation VI in scenarios where oblique RSFs were the underlying model used to compute VI and also in scenarios where other modeling techniques (\eg, boosted tree ensembles) were used.

@@

The current study has several limitations. The accelerated oblique RSF does not account for competing risks, and so our benchmark of prediction accuracy divided tasks where competing risks were present into event-specific tasks. Biased estimation of incidence may occur when competing risks are ignored, and allowing the oblique RSF to account for competing risks is a high priority for future studies. In addition, missing data are not addressed in the accelerated oblique RSF, and users are expected to impute missing values before passing training or testing data into exported functions from the \texttt{aorsf} R package. However, missing data are common and it is standard for ensemble tree methods to handle missing data during the tree growing procedure. Thus, a second item of high priority for future studies is to develop and evaluate strategies to handle missing data while growing an oblique RSF. Last, \citet{cui2017consistency} found that estimating an inverse-probability weighted hazard function at each non-leaf node of a survival tree allows the RSF to converge asymptotically to the true survival function when some variables contribute both to the risk of the event and the risk of censoring, a scenario that is very likely in the analysis of electronic medical records. The accelerated oblique RSF could incorporate this splitting technique by using Newton Raphson scoring to fit a model for the censoring distribution and then a weighted model could be fit to the failure distribution.

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\begin{landscape}
\begin{longtable}[t]{lcclcc}
\caption{\label{tab:unnamed-chunk-5}Data sets used for numeric experiments \label{tab:datasets}}\\
\toprule
Label & N observations & N predictors & Outcome & N Events & \% Censored\\
\midrule
VA lung cancer trial & 137 & 8 & Death & 128 & 6.57\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Recurrence & 468 & 49.6\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash Colon cancer} & \multirow{-2}{*}{\centering\arraybackslash 929} & \multirow{-2}{*}{\centering\arraybackslash 12} & Death & 452 & 51.3\\
\cmidrule{1-6}\pagebreak[0]
Primary biliary cholangitis & 276 & 19 & Death & 111 & 59.8\\
\cmidrule{1-6}\pagebreak[0]
Movies released in 2015-2018 & 551 & 46 & Gross 1M USD & 522 & 5.26\\
\cmidrule{1-6}\pagebreak[0]
GBSG II & 686 & 10 & Recurrence Or Death & 299 & 56.4\\
\cmidrule{1-6}\pagebreak[0]
Systolic Heart Failure & 2,231 & 41 & Death & 726 & 67.5\\
\cmidrule{1-6}\pagebreak[0]
Serum free light chain & 7,874 & 10 & Death & 2,169 & 72.5\\
\cmidrule{1-6}\pagebreak[0]
Non-alcohol fatty liver disease & 17,549 & 24 & Death & 1,364 & 92.2\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Recurrence & 1,518 & 49.1\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash Rotterdam tumor bank} & \multirow{-2}{*}{\centering\arraybackslash 2,982} & \multirow{-2}{*}{\centering\arraybackslash 11} & Death & 1,272 & 57.3\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & AIDS Diagnosis & 96 & 91.7\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash ACTG 320} & \multirow{-2}{*}{\centering\arraybackslash 1,151} & \multirow{-2}{*}{\centering\arraybackslash 12} & Death & 26 & 97.7\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Cardiovascular Death & 110 & 87.7\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash GUIDE-IT} & \multirow{-2}{*}{\centering\arraybackslash 894} & \multirow{-2}{*}{\centering\arraybackslash 59} & Hf Hospitalization & 288 & 67.8\\
\cmidrule{1-6}\pagebreak[0]
Early breast cancer & 614 & 1,692 & Recurrence Or Death & 134 & 78.2\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Cardiovascular Death & 521 & 94.4\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash SPRINT} & \multirow{-2}{*}{\centering\arraybackslash 9,361} & \multirow{-2}{*}{\centering\arraybackslash 174} & Death & 1,644 & 82.4\\
\cmidrule{1-6}\pagebreak[0]
Heart Transplant & 3,787 & 52 & Graft-Loss Or Death & 500 & 86.8\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Death & 76 & 86.0\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash FCL} & \multirow{-2}{*}{\centering\arraybackslash 541} & \multirow{-2}{*}{\centering\arraybackslash 7} & Relapse & 272 & 49.7\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Death & 963 & 30.4\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash Monoclonal gammopathy} & \multirow{-2}{*}{\centering\arraybackslash 1,384} & \multirow{-2}{*}{\centering\arraybackslash 8} & Malignancy & 115 & 91.7\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Heart Failure & 339 & 95.0\\
\cmidrule{4-6}\nopagebreak
 &  &  & Coronary Heart Disease & 439 & 93.5\\
\cmidrule{4-6}\nopagebreak
 &  &  & Stroke & 292 & 95.7\\
\cmidrule{4-6}\nopagebreak
\multirow{-4}{*}{\raggedright\arraybackslash MESA} & \multirow{-4}{*}{\centering\arraybackslash 6,783} & \multirow{-4}{*}{\centering\arraybackslash 48} & Death & 1,297 & 80.9\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Heart Failure & 2,981 & 78.1\\
\cmidrule{4-6}\nopagebreak
 &  &  & Coronary Heart Disease & 2,282 & 83.2\\
\cmidrule{4-6}\nopagebreak
 &  &  & Stroke & 1,323 & 90.3\\
\cmidrule{4-6}\nopagebreak
\multirow{-4}{*}{\raggedright\arraybackslash ARIC} & \multirow{-4}{*}{\centering\arraybackslash 13,623} & \multirow{-4}{*}{\centering\arraybackslash 41} & Death & 6,662 & 51.1\\
\bottomrule
\end{longtable}
\end{landscape}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\begin{longtable}[t]{lcclc}
\caption{\label{tab:unnamed-chunk-6}Index of prediction accuracy, time-dependent concordance statistic, and computational time required to fit and compute predictions for several learning algorithms across 31 risk prediction tasks.}\\
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Performance metric (SD)} & \multicolumn{2}{c}{Computation time, seconds} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
  & Scaled Brier & C-Statistic & Model fitting & Risk prediction\\
\midrule
\endfirsthead
\caption[]{Index of prediction accuracy, time-dependent concordance statistic, and computational time required to fit and compute predictions for several learning algorithms across 31 risk prediction tasks. \textit{(continued)}}\\
\toprule
  & Scaled Brier & C-Statistic & Model fitting & Risk prediction\\
\midrule
\endhead

@@

\hspace{1em}aorsf-fast & 0.134 (0.112) & 0.774 (0.071) & 0.772 & 0.188\\
\hspace{1em}aorsf-cph & 0.133 (0.112) & 0.773 (0.070) & 1.993 & 0.184\\
\hspace{1em}aorsf-net & 0.126 (0.127) & 0.770 (0.074) & 128.379 & 0.177\\
\hspace{1em}rsf-standard & 0.121 (0.116) & 0.760 (0.075) & 2.168 & 0.291\\
\hspace{1em}cif-standard & 0.120 (0.100) & 0.768 (0.070) & 4.511 & 15.781\\
\hspace{1em}cif-rotate & 0.120 (0.128) & 0.758 (0.080) & 42.206 & 9.111\\
\hspace{1em}obliqueRSF-net & 0.120 (0.086) & 0.773 (0.071) & 376.676 & 24.742\\
\hspace{1em}glmnet-cox & 0.117 (0.122) & 0.761 (0.075) & 0.460 & 0.004\\
\hspace{1em}ranger-extratrees & 0.105 (0.088) & 0.764 (0.067) & 2.416 & 1.775\\
\hspace{1em}cif-extension & 0.100 (0.095) & 0.763 (0.072) & 23.171 & 8.367\\
\hspace{1em}aorsf-random & 0.095 (0.083) & 0.743 (0.064) & 2.272 & 0.180\\
\hspace{1em}xgboost-cox & 0.072 (0.102) & 0.757 (0.088) & 4.368 & 0.004\\

>>>>

This study makes two contributions to oblique RSFs. First, we reduce their computational cost (\ie, accelerate them) with a scalable algorithm to identify linear combinations of coefficients. In a general benchmark experiment including 30 risk prediction tasks, we show that accelerated oblique RSFs are roughly 500 times faster with equivalent or superior prediction accuracy compared to existing routines to fit oblique RSFs. Second, we improve the interpretability of oblique RSFs with `negation VI', a method to estimate VI that flips the sign of coefficients in linear combinations of predictors instead of permuting predictor values. In simulation studies where VI is estimated using permutation, analysis of variance (ANOVA; see \citet{menze2011oblique}), and approximations of Shapley values, we find that negation VI improves the oblique RSF's ability to discriminate between correlated signal and noise variables. The accelerated oblique RSF and multiple methods to compute VI for oblique RSFs (permutation, ANOVA, and negation) are available in the \texttt{aorsf} R Package.

@@

Sections \ref{sec:bm_pred} and \ref{sec:bm_vi} present numerical experiments examining the accelerated oblique RSF and negation VI, respectively. The code used to run these experiments is available online at \href{https://github.com/bcjaeger/aorsf-bench}{https://github.com/bcjaeger/aorsf-bench}. All analyses were conducted using R version 4.1.3 with assistance from multiple R packages. To standardize comparisons of computational efficiency, all learners and VI techniques used up to 4 processing units.

@@

We use a collection of 17 data sets to benchmark the prediction accuracy and computational efficiency of the accelerated ORSF and each of the other learners described in \secref{sec:learners}. The number of right-censored outcomes per data set ranged from one to four, and the total number of risk prediction tasks we analyzed was 30 (\tabrefAppendix{tab:datasets}). Across all prediction tasks, the number of observations ranged from 137 to 17,549 (median: 1,807.5), the number of predictors ranged from 7 to 1,692 (median: 32.5), and the percentage of censored observations ranged from 5.26 to 97.7 (median: 78.1).

@@

For each risk prediction task, we completed 22 runs of Monte-Carlo cross validation. In each run, we used a random sample containing 50\% of the available data for training and the remaining 50\% for testing each of the learners described in \secref{sec:learners}. Then, for each learner, we computed the IPA, time-dependent C-statistic, and computational time required to fit a prediction model and compute risk predictions. If any learner failed to obtain predictions on any particular split of data\footnote{For example, when the prediction task was to predict risk of death in the ACTG 320 clinical trial (26 events total), some splits did not leave enough events in the training data to fit complex learners such as the neural network}, the results for that split were omitted from downstream analyses.

@@

After collecting data from 22 replications of Monte-Carlo cross validation for the 14 learners in all 30 risk prediction tasks, we analyzed the resulting 9,240 observations of IPA and, separately, time-dependent C-statistic, using a Bayesian linear mixed model. Our approach follows the ideas described by \citet{benavoli2017time} and \citet{tidymodels}, who developed guidelines on making statistical comparisons between learners using Bayesian models. Specifically, we fit two models: $$\text{IPA} = \widehat{\gamma}_0 + \widehat{\gamma} \cdot \text{learner} + (1\,|\, \text{data/run}) $$ and $$\text{C-stat} = \widehat{\gamma}_0 + \widehat{\gamma} \cdot \text{learner} + (1\,|\, \text{data/run}).$$ Random intercepts for specific splits of data (\ie, \texttt{run} in the model formula) were nested within datasets. The intercept, $\widehat{\gamma}_0$, was the expected value of the outcome using \texttt{aorsf-fast}, making the coefficients in $\widehat{\gamma}$ the expected differences between \texttt{aorsf-fast} and other learners. Default priors from \texttt{rstanarm} were applied for model fitting \citep{rstanarm}.

\paragraph{Hypothesis testing} For both the IPA and time-dependent C-statistic, we conducted equivalence and inferiority tests based on a 1 point region of practical equivalence. More specifically, we concluded that two learners had practically equivalent IPA or time-dependent C-statistic if there was a 95\% or higher posterior probability that the absolute difference in the relevant metric was less than 1. We concluded that one learner was weakly superior when there was $\geq$ 95\% posterior probability that the difference in the relevant metric was non-zero, and concluded superiority when when there was $\geq$ 95\% posterior probability that the difference in the relevant metric was 1 or more.

@@

Compared to learners that were not oblique RSFs, \texttt{aorsf-fast} had the highest IPA in 16 out of 30 risk prediction tasks, with an overall mean IPA of 13.6 (Figure \ref{fig:bm_pred_viz_ibs}). Compared to the learner with the second highest mean IPA (\texttt{rsf-standard}), \texttt{aorsf-fast}'s mean was 1.32 points higher, a relative increase of 10.7\%. The posterior probability of \texttt{aorsf-fast} and \texttt{aorsf-cph} having practically equivalent expected IPA was 0.97, and the posterior probability of \texttt{aorsf-fast} having a superior IPA to other learners ranged from 0.79 (versus \texttt{rsf-standard}) to $>$0.999 (versus several other learners; see Figure \ref{fig:bm_pred_model_viz_ibs})

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/bm_pred_viz_ibs-1} \caption[Index of prediction accuracy for the accelerated oblique random survival forest and other learning algorithms across multiple risk prediction tasks]{Index of prediction accuracy for the accelerated oblique random survival forest and other learning algorithms across multiple risk prediction tasks. Text appears in tasks where the accelerated oblique random survival forest obtained the highest index of prediction accuracy, showing the absolute and percent improvement over the second best learner. As predicted survival probabilities are not a standard output from \texttt{xgboost-aft}, it is not included in this figure.}\label{fig:bm_pred_viz_ibs}
\end{figure}

@@

Compared to learners that were not oblique RSFs, \texttt{aorsf-fast} had the highest time-dependent C-statistic in 10 out of 30 risk prediction tasks, with an overall mean of 77.6 (Figure \ref{fig:bm_pred_viz_cstat}). Compared to the learner with the second highest mean C-statistic (\texttt{cif-standard}), \texttt{aorsf-fast}'s mean was 0.664 points higher, a relative increase of 0.864\%. The posterior probability of \texttt{aorsf-fast} and \texttt{aorsf-cph} having practically equivalent expected time-dependent C-statistics was $>$ 0.999, and the posterior probability of \texttt{aorsf-fast} having a superior time-dependent C-statistic versus other learners ranged from 0.19 (versus \texttt{cif-standard}) to $>$0.999 (versus several other learners; see Figure \ref{fig:bm_pred_model_viz_cstat})

@@

Overall, \texttt{aorsf-fast} was the second fastest learner, with an expected model development and risk prediction time about 1/2 second longer than \texttt{glmnet-cox} (Figure \ref{fig:bm_pred_time}). Comparing median computing times, \texttt{aorsf-fast} was 525.2 times faster than its predecessor, \texttt{obliqueRSF-net}. In addition, \texttt{aorsf-fast} was 20.5, 4.27, and 2.79 faster than axis based forests grown using the \texttt{party}, \texttt{ranger}, and \texttt{randomForestSRC} packages, respectively.

@@

Parameters that varied in the current simulation study included the number of observations (500, 1000, and 2500) and the absolute value of the maximum correlation between predictors (0.3, 0.15, and 0). Parameters that remain fixed throughout the study included the number of predictors in each class (15) and the effect size of each predictor (one standard deviation increase associated with a 64\% increase in relative risk).

@@

The three techniques that used `aorsf' to estimate VI were ranked first (\texttt{aorsf-negate}; $C = 76.0$), second (\texttt{aorsf-anova}; $C = 74.0$), and third (\texttt{aorsf-permute}; $C = 73.3$) in overall mean C-statistic across all of the simulation scenarios, with \texttt{aorsf-negate} obtaining the highest C-statistic in 27 out of 36 VI tasks (Figure \ref{fig:bm_vi_viz}). Among the four relevant variable classes, \texttt{aorsf-negate} had the highest mean C-statistic for main effects, combination effects, and non-linear effects, with the greatest advantage of using \texttt{aorsf-negate} occurring among non-linear and combination variables. Full results from the experiment are provided in \tabrefAppendix{tab:bm_vi}. Computationally, ANOVA VI was faster than negation and permutation VI, with a median time of 1.34 seconds versus 12.3 and 14.1 seconds, respectively.

@@

In this paper, we have developed two contributions to the oblique RSF: (1) the accelerated oblique RSF (\ie, \texttt{aorsf-fast}) and (2) negation VI. Our technique to accelerate the oblique RSF reduces the number of operations required to find linear combinations of inputs using a single iteration of Newton Raphson scoring, while our VI technique directly engages with coefficients in linear combinations of inputs to measure importance of individual variables. In numeric experiments, we found that that \texttt{aorsf-fast} is over 500 times faster and just as accurate in risk prediction tasks compared to its predecessor, \texttt{obliqueRSF-net}. We also found that negation VI, a technique to estimate VI using the oblique RSF, detected non-linear, combination, and main effects more effectively than three standard methods to estimate VI: permutation, ANOVA, and SHAP VI. Overall, we found that estimating VI using negation instead of ANOVA increased the C-statistic for ranking a relevant variable higher than an irrelevant variable by 2.00, a relative increase of 2.70\%.

@@

While the current study advances the oblique RSF towards being scalable and interpretable, there remain several limitations that can be targeted in future studies. The accelerated oblique RSF does not account for competing risks, and biased estimation of incidence may occur when competing risks are ignored. Thus, allowing the oblique RSF to account for competing risks is a high priority for future studies. In addition, missing data are not addressed in the accelerated oblique RSF, and users of the \texttt{aorsf} R package are expected to impute missing values prior to model training and testing. However, missing data are common and there are numerous techniques for ensemble tree methods to handle missing data during the tree growing procedure. Thus, a second item of high priority for future studies is to develop and evaluate strategies to handle missing data while growing an oblique RSF. Last, \citet{cui2017consistency} found that estimating an inverse-probability weighted hazard function at each non-leaf node of a survival tree allows the RSF to converge asymptotically to the true survival function when some variables contribute both to the risk of the event and the risk of censoring, a scenario that is very likely in the analysis of electronic medical records. The accelerated oblique RSF could incorporate this splitting technique by using Newton Raphson scoring to fit a model for the censoring distribution and then a weighted model could be fit to the failure distribution. This final item has the highest priority, as \citet{cui2017consistency} showed it is a requisite condition for consistency of axis based survival trees in fairly general settings.

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\begin{landscape}
\begin{longtable}[t]{lcclcc}
\caption{\label{tab:unnamed-chunk-8}Data sets used for numeric experiments \label{tab:datasets}}\\
\toprule
Label & N observations & N predictors & Outcome & N Events & \% Censored\\
\midrule
VA lung cancer trial & 137 & 8 & Death & 128 & 6.57\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Recurrence & 468 & 49.6\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash Colon cancer} & \multirow{-2}{*}{\centering\arraybackslash 929} & \multirow{-2}{*}{\centering\arraybackslash 12} & Death & 452 & 51.3\\
\cmidrule{1-6}\pagebreak[0]
Primary biliary cholangitis & 276 & 19 & Death & 111 & 59.8\\
\cmidrule{1-6}\pagebreak[0]
Movies released in 2015-2018 & 551 & 46 & Gross 1M USD & 522 & 5.26\\
\cmidrule{1-6}\pagebreak[0]
GBSG II & 686 & 10 & Recurrence Or Death & 299 & 56.4\\
\cmidrule{1-6}\pagebreak[0]
Systolic Heart Failure & 2,231 & 41 & Death & 726 & 67.5\\
\cmidrule{1-6}\pagebreak[0]
Serum free light chain & 7,874 & 10 & Death & 2,169 & 72.5\\
\cmidrule{1-6}\pagebreak[0]
Non-alcohol fatty liver disease & 17,549 & 24 & Death & 1,364 & 92.2\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Recurrence & 1,518 & 49.1\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash Rotterdam tumor bank} & \multirow{-2}{*}{\centering\arraybackslash 2,982} & \multirow{-2}{*}{\centering\arraybackslash 11} & Death & 1,272 & 57.3\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & AIDS Diagnosis & 96 & 91.7\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash ACTG 320} & \multirow{-2}{*}{\centering\arraybackslash 1,151} & \multirow{-2}{*}{\centering\arraybackslash 12} & Death & 26 & 97.7\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Cardiovascular Death & 110 & 87.7\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash GUIDE-IT} & \multirow{-2}{*}{\centering\arraybackslash 894} & \multirow{-2}{*}{\centering\arraybackslash 59} & Hf Hospitalization & 288 & 67.8\\
\cmidrule{1-6}\pagebreak[0]
Early breast cancer & 614 & 1,692 & Recurrence Or Death & 134 & 78.2\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Cardiovascular Death & 521 & 94.4\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash SPRINT} & \multirow{-2}{*}{\centering\arraybackslash 9,361} & \multirow{-2}{*}{\centering\arraybackslash 174} & Death & 1,644 & 82.4\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Death & 76 & 86.0\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash FCL} & \multirow{-2}{*}{\centering\arraybackslash 541} & \multirow{-2}{*}{\centering\arraybackslash 7} & Relapse & 272 & 49.7\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Death & 963 & 30.4\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash Monoclonal gammopathy} & \multirow{-2}{*}{\centering\arraybackslash 1,384} & \multirow{-2}{*}{\centering\arraybackslash 8} & Malignancy & 115 & 91.7\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Heart Failure & 339 & 95.0\\
\cmidrule{4-6}\nopagebreak
 &  &  & Coronary Heart Disease & 439 & 93.5\\
\cmidrule{4-6}\nopagebreak
 &  &  & Stroke & 292 & 95.7\\
\cmidrule{4-6}\nopagebreak
\multirow{-4}{*}{\raggedright\arraybackslash MESA} & \multirow{-4}{*}{\centering\arraybackslash 6,783} & \multirow{-4}{*}{\centering\arraybackslash 48} & Death & 1,297 & 80.9\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Heart Failure & 2,981 & 78.1\\
\cmidrule{4-6}\nopagebreak
 &  &  & Coronary Heart Disease & 2,282 & 83.2\\
\cmidrule{4-6}\nopagebreak
 &  &  & Stroke & 1,323 & 90.3\\
\cmidrule{4-6}\nopagebreak
\multirow{-4}{*}{\raggedright\arraybackslash ARIC} & \multirow{-4}{*}{\centering\arraybackslash 13,623} & \multirow{-4}{*}{\centering\arraybackslash 41} & Death & 6,662 & 51.1\\
\bottomrule
\end{longtable}
\end{landscape}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\begin{longtable}[t]{lcccc}
\caption{\label{tab:unnamed-chunk-9}Index of prediction accuracy, time-dependent concordance statistic, and computational time required to fit and compute predictions for several learning algorithms across 31 risk prediction tasks.\label{tab:bm_pred_all}}\\
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Performance metric (SD)} & \multicolumn{2}{c}{Computation time, seconds} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
  & Scaled Brier & C-Statistic & Model fitting & Risk prediction\\
\midrule
\endfirsthead
\caption[]{Index of prediction accuracy, time-dependent concordance statistic, and computational time required to fit and compute predictions for several learning algorithms across 31 risk prediction tasks.\label{tab:bm_pred_all} \textit{(continued)}}\\
\toprule
  & Scaled Brier & C-Statistic & Model fitting & Risk prediction\\
\midrule
\endhead

@@

\hspace{1em}aorsf-fast & 0.136 (0.113) & 0.776 (0.072) & 0.721 & 0.174\\
\hspace{1em}aorsf-cph & 0.136 (0.113) & 0.775 (0.071) & 1.867 & 0.169\\
\hspace{1em}aorsf-net & 0.129 (0.128) & 0.771 (0.075) & 131.131 & 0.167\\
\hspace{1em}rsf-standard & 0.123 (0.118) & 0.761 (0.076) & 2.116 & 0.278\\
\hspace{1em}cif-standard & 0.123 (0.101) & 0.769 (0.071) & 4.312 & 14.590\\
\hspace{1em}obliqueRSF-net & 0.122 (0.087) & 0.775 (0.072) & 379.917 & 23.183\\
\hspace{1em}cif-rotate & 0.121 (0.130) & 0.759 (0.082) & 39.895 & 8.598\\
\hspace{1em}glmnet-cox & 0.119 (0.124) & 0.763 (0.075) & 0.433 & 0.004\\
\hspace{1em}ranger-extratrees & 0.107 (0.088) & 0.765 (0.067) & 1.986 & 1.289\\
\hspace{1em}cif-extension & 0.102 (0.096) & 0.764 (0.072) & 22.524 & 7.926\\
\hspace{1em}aorsf-random & 0.097 (0.084) & 0.744 (0.064) & 2.201 & 0.169\\
\hspace{1em}xgboost-cox & 0.074 (0.104) & 0.759 (0.089) & 4.384 & 0.004\\
\hspace{1em}nn-cox & 0.050 (0.109) & 0.652 (0.138) & 16.298 & 3.578\\
